---
output: md_document
---

    ---
    layout: post
    title: "Using the `afex` R package for ANOVA (factorial and repeated measures)"
    categories: jekyll update
    author: Tyson S. Barrett
    comments: true
    ---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We recently switched our graduate statistics courses to `R` from SPSS (yay!). It has gone fairly well. However, once we get into ANOVA-type methods, particularly the repeated measures flavor of ANOVA, `R` isn't as seamless as almost every other statistical approach. As such, my colleague [Sarah Schwartz](sarahschwartzstats.com) found the [`afex`](https://github.com/singmann/afex) package that looks like it can be helpful in simplifying the code and increasing the amount of useful information obtained.

This post is for walking through the use of `afex`, includig `aov_car()`, `aov_ez()`, and `aov_4()`. To show them off, I'll use the following ficticious data set:

```{r}
set.seed(42)
z <- data.frame(a1 = c(rnorm(100,2), rnorm(100,1),rnorm(100,0)),
                b  = rep(c("A", "B", "C"), each = 100),
                c  = factor(rbinom(300, 1, .5)),
                ID = 1:300,
                a2 = c(rnorm(100,2), rnorm(100,1),rnorm(100,0)),
                a3 = c(rnorm(100,2), rnorm(100,1),rnorm(100,0)))
```

and we will load:

```{r, message = FALSE, warning = FALSE}
library(lsmeans)
library(afex)
library(tidyverse)
```


## One-Way ANOVA
#### `afex::aov_car()`

The main function in `afex` is `aov_car()`. Both of the functions described in the later sections are just wrappers of this one but allow different syntax. To use `aov_car()`, we are essentially using aov with a few adjustments.
```{r}
aov1 <- z %>%
  aov_car(a1 ~ b + Error(ID),
          data = .)
aov1
```
Notably, we use the `Error(ID)` even in one-way ANOVA. This complication can actually simplify things, as changing to a repeated-measures or mixed effects model is not difficult. Essentially all it is doing is telling the function the grouping variable.

We can check assumptions fairly quickly with the `plot()` function and pulling out the `aov` object from `aov1`.

```{r}
plot(aov1$aov)
```


From here, we can obtain the least squares means and plot them with `ggplot2` showing the mean and the confidence interval. It is important to not pull the `lsmeans()` function from the `lsmeans` package because it actually is functionality that `afex` provides. As such, using `lsmeans::lsmeans()` will through an error.
```{r}
aov1 %>%
  lsmeans(specs = "b") %>%
  data.frame %>%
  ggplot(aes(b, lsmean)) +
    geom_point() +
    geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL))
```

We can also look at some PostHoc type analyses using the `lsmeans()` object.

```{r}
aov1 %>%
  lsmeans(specs = "b") %>%
  pairs() %>%
  update(by=NULL, adjust = "tukey")
```
We can also use "holm", "bonf", and "scheffe" in place of "tukey". In this case, it doesn't matter much because they have very small p-values.


#### `afex::aov_ez()`

This function is notable in its explicit syntax; however, my preferences are for `aov_4()` for most situations because it uses the syntax of `lme4::lmer()`.

```{r}
z %>%
  aov_ez(id = "ID",
         dv = "a1",
         between = "b",
         data = .)
```

The same checks and tests as shown before are possible here.


#### `afex::aov_4()`

```{r}
z %>%
  aov_4(a1 ~ b + (1|ID),
        data = .)
```

Again, the same checks and tests can be done with this one as well.


## Factorial ANOVA

Two or more factors can be included as well. These are very similar to their One-Way ANOVA counterparts.

```{r}
aov2 <- z %>%
  aov_car(a1 ~ b * c + Error(ID),
          data = .)

aov2.1 <- z %>%
  select(a1, b, c, ID) %>%
  aov_ez(id = "ID",
         dv = "a1",
         between = c("b", "c"),
         data = .)

aov2 <- z %>%
  aov_4(a1 ~ b * c + (1|ID),
        data = .)
aov2
```
You can remove the interaction in the following ways:
```{r}
aov2.1 <- z %>%
  aov_car(a1 ~ b + c + Error(ID),
          data = .)

## Not sure how to remove it from `aov_ez()`

aov2.1 <- z %>%
  aov_4(a1 ~ b + c + (1|ID),
        data = .)
aov2.1
```



We can check some assumptions using the `plot()` function.

```{r}
aov2$aov %>%
  plot()
```


From here, we can obtain the least squares means and plot them with `ggplot2` showing the mean and the confidence interval by our other factor.
```{r}
aov2 %>%
  lsmeans(specs = c("b", "c")) %>%
  data.frame %>%
  ggplot(aes(b, lsmean, group = c, color = c)) +
    geom_point(position = position_dodge(width = .3)) +
    geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL),
                  position = position_dodge(width = .3),
                  width = .2)
```


### Repeated Measures ANOVA

Repeated measures is now much like the other types of ANOVA. The major difference is that we need to change the format of the data.

```{r}
z_long <- z %>%
  tidyr::gather("meas", "value", a1, a2, a3)
```

```{r, echo=FALSE}
as.tibble(z_long)
```

```{r}
aov_rm <- z_long %>%
  aov_car(value ~ 1 + Error(ID/meas),
          data = .)
aov_rm
```
This, however, is when the `aov_4()` is easier for me to understand overall. It uses code that makes more sense to me so I'll show it here (I won't be showing how to use `aov_ez()` for repeated measures).

```{r}
aov_rm <- z_long %>%
  aov_4(value ~ 1 + (meas|ID),
        data = .)
aov_rm
```

Checking assumptions with the repeated measures ANOVA is notably harder, in general and in `R`. Here, we are waiting for some development to happen so we can do the following (note the use of `lm` instead of `aov`).
```{r, eval=FALSE}
aov_rm$lm %>%
  plot()
```

    Error: 'plot.mlm' is not implemented yet

We can obtain the least squares means and plot them with `ggplot2` showing the mean and the confidence interval.
```{r}
aov_rm %>%
  lsmeans(specs = "meas") %>%
  data.frame %>%
  ggplot(aes(meas, lsmean)) +
    geom_point() +
    geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL))
```


## Mixed Models

Similarly to the repeated measures ANOVA, I'm going to focus on `aov_4()`.

```{r}
mixed_mod <- z_long %>%
  aov_4(value ~ b + c + (meas|ID),
        data = .)
mixed_mod
```

We have the same issue of checking assumptions here as the repeated measures ANOVA
```{r, eval=FALSE}
mixed_mod$lm %>%
  plot()
```
      
      Error: 'plot.mlm' is not implemented yet

Finally, we can check out the least-squares means across groups and time.

```{r}
mixed_mod %>%
  lsmeans(specs = c("meas", "b", "c")) %>%
  data.frame %>%
  ggplot(aes(meas, lsmean, group = b, color = b)) +
    geom_point() +
    geom_line() +
    geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL),
                  width = .2) +
    facet_grid(~c)
```



There you have it! I believe `afex` is a valuable contribution to `R`, particularly in dealing with factorial and repeated-measures ANOVA.



