---
title: "dtplyr intro"
author: "Tyson S. Barrett"
date: "10/14/2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


So you want to learn `dtplyr`? Well, you are in luck if you have used `dplyr`.

```{r}
library(data.table)
library(dtplyr)
library(dplyr)
```

We will start with the following data table.

```{r}
dt <- data.table(
  x = rnorm(1e6),
  y = runif(1e6),
  grp = sample(1L:3L, 1e6, replace = TRUE)
)
dt
```

In order to use the `dplyr` verbs on a data table (while using the `data.table` packages strengths), we need to create a "lazy data table". 

```{r}
ldt <- lazy_dt(dt)
```

Then we can simply use the core `dplyr` verbs. For example, we can do the following.

```{r}
ldt %>% 
  mutate(z = rnorm(1e6)) %>% 
  filter(grp == 1) %>% 
  show_query()
```

By ending with `show_query()` we get to see what the pipeline will call, in this case, ```copy(`_DT1`)[, `:=`(z = rnorm(1e+06))][grp == 1]```. To actually have it perform the call, we can use `as.data.table()`, `as.data.frame()`, or `as_tibble()`.

```{r}
ldt2 <- ldt %>% 
  mutate(z = rnorm(1e6)) %>% 
  filter(grp == 1) %>% 
  as.data.table()
ldt2
```

Without the `as.data.table()` in this example, we'd get the following warning from `dplyr`:

    Warning message:
    You are using a dplyr method on a raw data.table, which will call the data frame implementation, and is likely to be inefficient.

    To suppress this message, either generate a data.table translation with `lazy_dt()` or convert to a data frame or tibble with `as.data.frame()`/`as_tibble()`.

This means, to use `dtplyr`, we really only have two additional steps:

1. `lazy_dt()`
2. `as.data.table()`

That is all. Everything else is `dplyr`.


## Combining `data.table` and `dtplyr`

The current edition of `dtplyr` is still very new, and as such, is not going to provide the vast functionality that `dplyr` has. This means we'll often need to use `data.table` syntax directly or use raw `dplyr`. Let's say we have two tables, one with demographic information, another with response information for a study we are conducting. These tables could look like the following:

```{r, echo = FALSE}
dt_1 <- ldt %>% 
  mutate(id = sample(1L:1e6L)) %>% 
  mutate(icc = rnorm(1e6, id/1e6)) %>% 
  mutate(outcome1 = x + y + grp + icc + rnorm(1e6),
         outcome2 = outcome1 + grp + icc + rnorm(1e6)) %>% 
  select(-icc) %>% 
  as.data.table()

dt_2 <- data.table(
  id = sample(1L:1e6L),
  sex = sample(1L:3L, 1e6, replace = TRUE),
  location = sample(c("urban", "rural", "suburbia", "nowhere"), 1e6, replace = TRUE),
  income = sample(40000:90000, 1e6, replace = TRUE)
)
```

```{r}
dt_1
dt_2
```

where, in `dt_1`, `x` and `y` are measured variables, `grp` was assigned, `id` is unique for each individual, and `outcome1` and `outcome2` are outcome variables for time 1 and 2, respectively; in `dt_2` the `id` matches the one in `dt_1`, and `sex`, `location`, and `income` are variables per individual.

To begin working with this data, we might want to join the two sets, reshape them, clean (if necessary), and then analyze it.


### Step 1. Joining

We will do a full join below. But first, we need to create "lazy dt" tables and then don't forget to call `as.data.table()`.

```{r}
dt_1 <- lazy_dt(dt_1)
dt_2 <- lazy_dt(dt_2)

full_dt <- full_join(dt_1, dt_2, by = "id") %>% 
  as.data.table()
full_dt
```

The `full_join(dt_1, dt_2, by = "id")` makes a call to the `merge.data.table()` method.

### Step 2. Reshaping

We can then reshape it to be in "tidy" format. We'll use the `melt` function in `data.table` for this, given its efficiency and speed.

```{r}
melted <- melt(full_dt, measure.vars = c("outcome1", "outcome2"), id.vars = c("id", "sex", "location", "income"))
```

By ordering by `id` we can see that each individual has two rows, one for `outcome1` and another for `outcome2`.

```{r}
melted[order(id)]
```


### Step 3. Cleaning

We may also want to put labels on the `sex` variable. We can do this with:

```{r}
melted_clean <- 
  melted %>% 
  lazy_dt() %>% 
  mutate(sex = factor(sex,
                      labels = c("male", "female", "non-binary"))) %>% 
  as.data.table()
melted_clean
```


### Step 4. Analyzing

Finally, we can compute descriptives or run models with these data. For example, we can run:

```{r}
melted_clean[, .(mean = mean(value)), by = variable]
```

or

```{r}
melted_clean[, .(mean = mean(income)), by = .(location)]
```










