---
title: "Nesting with dtplyr and tidyfast"
author: "Tyson S. Barrett"
date: "11/13/2019"
output: md_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options("width" = 150)
```


With the advent of a more cohesive and complete `dtplyr`, I've been wanting to write about how it can be used with `tidyfast` to use the syntax of the `tidyverse` while relying on the speed and efficiency of `data.table`. This workflow is already being adopted by some, including [Ivan Leung](https://twitter.com/urganmax?ref_src=twsrc%5Etfw%7Ctwcamp%5Etweetembed%7Ctwterm%5E1193410459745038336&ref_url=https%3A%2F%2Fpublish.twitter.com%2F%3Fquery%3Dhttps%253A%252F%252Ftwitter.com%252Furganmax%252Fstatus%252F1193410459745038336%26widget%3DTweet), who posted:

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">Refactoring an existing nested+mutate+map() with {tidyfast} equivalent has cut data processing time from 2.5 hrs to 30 min 🙇🏻
♂️🙏🏼</p>&mdash; Ivan Leung (@urganmax) <a href="https://twitter.com/urganmax/status/1193410459745038336?ref_src=twsrc%5Etfw">November 10, 2019</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

This is the first in this series of `dtplyr` and `tidyfast` posts---this one regarding nesting and unnesting data table into and from list-columns. This post uses `tidyfast` version 0.1.8 and `dtplyr` version 1.0.0.

## TL;DR

`dtplyr` is an intutitive approach to using `data.table` while maintaining syntax from `dplyr`. It translates `dplyr` code to `data.table` and doesn't evaluate the call until the user calls one of several functions (e.g. `as.data.table()`, `as_tibble()`, or `as.data.frame()`). The structure of using `dtplyr` can be summarized as:

```{r, eval = FALSE}
lazy_dt(data) %>% 
  <dplyr code> %>% 
  <dplyr code> %>% 
  as.data.table()
```

where `<dplyr code>` is any of the supported `dplyr` functions. This, particularly in larger data sets, has profound speed and, at times, efficiency improvements.

This general approach, that of using `data.table` "under the hood" while using the `tidyverse`s explicit grammar of data manipulation, has been of interest to me over the past several months. It just happened to be around the time that Hadley Wickham picked up `dtplyr` again and reframed it in a way that allows `data.table` to do the heavy data lifting while still maintaining all the flexbility of the `tidyverse`. 

Herein, it is also shown how `tidyfast` can be used in conjunction with `dtplyr` to nest and unnest data tables. This allows for a seamless *nest-map-analyze* type workflow.


## Intro to `dtplyr`

The `dtplyr` package actually only has one function: `lazy_dt()`. This creates a speacial type of `data.table` that, in essence, records any `dplyr` call and saves it until we use something like `as.data.table()`---at which time the call is actually evaluated.

Let's use the `flights` data set from the `nycflights13` package.

```{r, message=FALSE, warning=FALSE}
library(dtplyr)
library(dplyr)
library(data.table)
library(tidyfast)

starwars <- dplyr::starwars
```

These data have `r nrow(starwars)` rows and `r ncol(starwars)` columns, a great situation (although many more columns and rows can certainly be handled) for some speedy computations.

Before starting any examples, in `tidyfast` there is a `dt_print_options()` function that adjusts the default printing options with `data.table`.

```{r}
dt_print_options()
```


For a simple example, we can filter some rows, add a new variable, and select columns. 

```{r}
starwars2 <- starwars %>% 
  lazy_dt() %>%                                 # create the lazy_dt
  filter(species == "Human") %>%                # only humans
  mutate(height_inch = height*0.3937) %>%       # cm to inches
  select(name, height_inch, homeworld,          # select some vars
         species)
```

Since we haven't called `as.data.table()` or `as_tibble()`, there haven't been any actual evaulated calls yet. We can see what the `data.table` call looks like with:

```{r}
show_query(starwars2)
```

To have it actually call, let's save it as a `data.table` and then print it (but only a few of the variables).

```{r}
starwars2 <- as.data.table(starwars2)
starwars2[]                           # the [] forces it to print
```


## Integrating `tidyfast` and `dtplyr`

`tidyfast` can help in a few areas that `dtplyr` doesn't cover (as of right now). The first, and the emphasis of this post, is with nesting and unnesting data into and from list-columns of data tables. A [recent pre-print](https://psyarxiv.com/u8ekc/) goes into more detail on the *why* of nesting/unnesting in your workflow (while introducing some basic code that `tidyfast::dt_nest()` and `tidyfast::dt_unnest()` were built on). It states in the abstract:

> "[Using nesting provides] a cognitively efficient way to organize results of complex data (e.g. several statistical models, groupings of text, data summaries, or even graphics) with corresponding data. For example, one can store student information within classrooms, player information within teams, or even analyses within groups."

To take advantage of this data analytic approach using `data.table`, `dtplyr`, and `tidyfast`, the user can use the following general workflow:

```{r, eval = FALSE}
nested <- lazy_dt(data, immutable = FALSE) %>% 
  <dplyr code> %>% 
  <dplyr code> %>% 
  dt_nest(...)
```

where `...` are groups to nest by. Since `dt_nest()` automatically calls `as.data.table()` on anything other than a `data.table` object (in this case the `lazy_dt`), it will evaluate the `dtplyr` calls and then do the nesting all in that one line. Notably, the `immutable = FALSE` makes it so `dtplyr` isn't making copies of the data. This becomes helpful when it comes to working with nesting/unnesting.

After doing all the necessary nesting data manipulations and analyses, we can then unnest with `dt_unnest()`.

### A Short Example

For example, we may want to analyze the `starwars` data by movie. The issue here is that the films are already in a list-column, as shown below:

```{r}
select(starwars, name, species, homeworld, films)
```

So we need to unnest that column and then nest by each film. To do so, we are going to use `dt_hoist()` and then `dt_nest()`.

```{r}
films_dt <- dt_hoist(starwars,
                     films)
films_dt

films_nest <- dt_nest(films_dt, films)
films_nest
```

So `films_nest` has two columns, one is the film and the second is the data that goes with each film. This allows us to see the movies that are included in this data set and organizes the data in such a way to help us work with it within each film more safely.

Let's start using `dtplyr` here. Below, we find the number of homeworlds represented in each movie and return a double vector (using `map_dbl()`) and get the counts of each gender represented in each movie, returning a `data.table`.


```{r}
films_nest <- films_nest %>% 
  lazy_dt() %>% 
  mutate(num_worlds = purrr::map_dbl(data, ~length(unique(.x$homeworld)))) %>%   
  mutate(prop_genders = purrr::map(data, ~dt_count(.x, gender))) %>% 
  as.data.table()
films_nest[]
```

Turns out we also have how much each movie grossed at the box office (according to [Statista](https://www.statista.com/statistics/311356/star-wars-production-costs-box-office-revenue/))

```{r}
revenue <- 
  tibble::tribble(
    ~films,                   ~revenue,
    "A New Hope",                775.4,
    "The Empire Strikes Back",   538.4,
    "Return of the Jedi",        475.1,
    "The Phantom Menace",       1027.0,
    "Attack of the Clones",      649.4,
    "Revenge of the Sith",       848.8,
    "The Force Awakens",        2068.2
  ) %>% 
  as.data.table()

films_nest <- films_nest[revenue, on = "films"]
films_nest
```

With this information, we can extract the proportion of females represented in the movie and the box office revenue. We can do this in a few steps. First, let's unnest the `prop_genders` variable. We will use `dt_unnest()` as it is designed to unnest columns with data tables. Notice that the other list-column (`data`) is dropped from this operation. This is partly a safety feature to avoid copying what could already be large data in the list-columns.

```{r}
films_unnest <- dt_unnest(films_nest, prop_genders)
films_unnest
```

Using this, we can calculate, per film, the proportion of females represented.

```{r}
prop_female <- films_unnest %>% 
  lazy_dt() %>% 
  group_by(films) %>% 
  mutate(sum_characters = sum(N)) %>% 
  filter(gender == "female") %>% 
  mutate(prop_female = N/sum_characters) %>% 
  as.data.table()
prop_female[]
```

With this info, we visualize how these relate.

```{r, fig.height=4, fig.width=5, dpi=200}
library(ggplot2)
# set the theme
theme_set(
  theme_minimal() +
  theme(legend.position = "none",
        panel.border = element_rect(color = "lightgrey", 
                                    fill = NA),
        panel.background = element_rect(fill = "#FBFCFC"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linetype = "dotted"))
)

prop_female %>% 
  mutate(revenue = as.numeric(revenue)) %>% 
  ggplot(aes(prop_female, revenue, color = films)) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = films),
                           nudge_y = 50) +
  labs(x = "Proportion of Major Female Characters",
       y = "Box Office Revenue",
       caption = "Data from Statista and SWAPI.") +
  scale_color_viridis_d(end = .92)
```

Looks like, at least in this very small sample, that more females that are major characters in the movies is somewhat positively related to box office earnings. This is largely driven by *The Force Awakens*. Without that one, it would look like there wasn't really much of a relationship between the two.


## Conclusion

This was a brief introduction to `dtplyr` and how it works with some `tidyfast` functions. Hopefully the workflow presented herein will be helpful.

```{r}
sessioninfo::session_info()
```


