---
title: "NBA Player Stats"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

## Wrangle That NBA Data

We will use the `rvest` package to download the data from [NBA Stuffer](nbastuffer.com), and `dplyr`, `data.table`, and `janitor` for data wrangling.
```{r}
library(rvest)
library(dplyr)
library(data.table)
library(janitor)
```

```{r}
players_2018 <- read_html("https://www.nbastuffer.com/2017-2018-nba-player-stats/")
players_2019 <- read_html("https://www.nbastuffer.com/2018-2019-nba-player-stats/")
team_2019 <- read_html("https://www.nbastuffer.com/2018-2019-nba-team-stats/")
team_2018 <- read_html("https://www.nbastuffer.com/2017-2018-nba-team-stats/")
```

Functions to help us extract the tables
```{r}
extract_fun <- function(html){
  html_nodes(html, "table") %>% 
    .[2] %>% 
    html_table(fill = TRUE) %>% 
    .[[1]] 
}
```


```{r}
player_2018 <- 
  extract_fun(players_2018) %>% 
  mutate(year = 2018,
         AGE = as.numeric(AGE))
player_2019 <- 
  extract_fun(players_2019) %>% 
  mutate(year = 2019)
players <- 
  bind_rows(player_2018, player_2019) %>% 
  clean_names() %>% 
  data.table()

team_2018 <- 
  extract_fun(team_2018) %>% 
  mutate(year = 2018)
team_2019 <- 
  extract_fun(team_2019) %>% 
  mutate(year = 2019)
team <- 
  bind_rows(team_2018, team_2019) %>% 
  clean_names() %>% 
  data.table()
```

## Improvement 2018 to 2019 Seasons

### PPG

The first metric we will look at is points per game.

```{r}
library(ggplot2)
library(ggrepel)
```

```{r}
theme_set(theme_minimal() +
          theme(legend.position = "none"))
```


```{r}
ppg <-
  players %>% 
  dcast(full_name ~ year, 
        value.var = "ppg_points_points_per_game", 
        fill = NA, 
        fun=mean) %>% 
  mutate(diff = `2019` - `2018`,
         extreme = case_when((diff > quantile(diff, .94, na.rm = TRUE) & `2019` > 12) |
                              (diff < quantile(diff, .02, na.rm = TRUE)) ~ 1,
                             TRUE ~ 0))
ggplot(ppg, aes(`2018`, `2019`, group = full_name)) +
  geom_abline(slope = 1) +
  geom_line(alpha = .5) +
  geom_point(point = 21,
             aes(alpha = extreme)) +
  geom_text_repel(data = filter(ppg, extreme == 1),
                  aes(label = full_name)) +
  labs(x = "2017-2018 Season", y = "2018-2019 Season")
```

### Assists Per Game

```{r}
apg <-
  players %>% 
  dcast(full_name ~ year, 
        value.var = "apg_assists_assists_per_game", 
        fill = NA, 
        fun=mean) %>% 
  mutate(diff = `2019` - `2018`,
         extreme = case_when((diff > quantile(diff, .94, na.rm = TRUE) & `2019` > 4) |
                              (diff < quantile(diff, .02, na.rm = TRUE)) ~ 1,
                             TRUE ~ 0))
ggplot(apg, aes(`2018`, `2019`, group = full_name)) +
  geom_abline(slope = 1) +
  geom_line(alpha = .5) +
  geom_point(point = 21,
             aes(alpha = extreme)) +
  geom_text_repel(data = filter(apg, extreme == 1),
                  aes(label = full_name)) +
  labs(x = "2017-2018 Season", y = "2018-2019 Season")
```

## Organizing Players by Team

Here, we will use the `data.table` package with the following custom function to create list columns (nested data).


```{r, eval=FALSE}
group_nest_dt <- function(dt, by = id, .key = "data"){
  stopifnot(is.data.table(dt))

  by <- substitute(list(by))
  
  express <- dt[, list(list(.SD)), by = eval(by)]
  setnames(express, old = "V1", new = .key)
  express
}

players[, .(team, full_name, mpg)] %>% 
  group_nest_dt(team) %>% 
  .$data

group_nest(players, team)

## Find a way to make it a data.frame at the beginning so we aren't creating a list then a data.frame

nested_dt <-
  nest.data.table(players, 
                  full_name, mpg, ppg_points_points_per_game, 
                  rpg_rebounds_rebounds_per_game, apg_assists_assists_per_game, 
                  bpg_blocks_blocks_per_game,
                  by = team)
nested_dt

## remotes::install_github("hadley/lineprof")
lineprof::lineprof(nest.data.table(players, 
                                    full_name, mpg, ppg_points_points_per_game, 
                                    rpg_rebounds_rebounds_per_game, apg_assists_assists_per_game, 
                                    bpg_blocks_blocks_per_game,
                                    by = team)) %>% 
  lineprof::shine()

test_set <- players %>% 
  data.frame() %>% 
  select(full_name, mpg, ppg_points_points_per_game, 
            rpg_rebounds_rebounds_per_game, apg_assists_assists_per_game, 
            bpg_blocks_blocks_per_game, team)

first = bench::mark(group_nest(players, team))
second = bench::mark(nest.data.table(players,by = team))

```




