---
title: "Untitled"
author: "Tyson S. Barrett, PhD"
date: "10/23/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```


# ANOVA and Friends in `R`

`R`, in general, makes most modeling procedures more straightforward, in my opinion. However, when it comes to ANOVA, it can be very confusing for students and researchers. To aleviate this, the `afex` package is designed to make many things easier. This post is designed to help you understand how to run ANOVAs in `R` using `afex`.

To start, let's load a few packages that will come in handy, including `afex`. Note that you will likely get messages from these packages. It is alright to ignore them.

```{r}
library(afex)
library(emmeans)
library(tidyverse)
library(furniture)
```

And we'll start with a data set from the `educ6600` package that can be installed with `remotes::install_github("tysonstanley/educ6600")` (you may need to install `remotes` first the normal way). This data set, called "data_wait", is then transformed to a long format (it was originally a wide format), with the grouping variable being `caller_type` and `wait_time` being the outcome. We then add an ID variable.

```{r}
data_wait_long <- educ6600::data_wait %>% 
  gather("caller_type", "wait_time") %>% 
  mutate(id = 1:n())
data_wait_long
```


## One-Way ANOVA

Below is a one way anova, that we assign to the name of `oneway`.

```{r}
oneway <- aov_4(wait_time ~ caller_type + (1 | id), data = data_wait_long)
```

This `oneway` object has several things inside:

```{r}
names(oneway)
```

By default, if we just print it out, we get a simple summary of the ANOVA results. This includes the F-statistic, the global effect size (`ges`), and the p-value.

```{r}
oneway
```















