---
output: pdf_document
header-includes: 
   \usepackage{graphicx}
   \usepackage{fancyhdr}
   \pagestyle{fancy}
   \setlength\headheight{28pt}
   \fancyhead[L]{Tyson S. Barrett}
   \fancyhead[R]{MLM}
   \usepackage{ebgaramond}
   \usepackage[ugm]{mathdesign}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Multilevel Analysis by *Joop J. Hox*
## Chapter 5 [^start]

The chapter discusses how to analyze the special multilevel case of *longitudinal* data. There are two main types: 1) fixed occasions and 2) varying occasions. Multilevel modeling (MLM) is flexible enough to handle both (whereas the ANOVA family of methods can only handle fixed). MLM does not remove an individual if they have missing values, but rather do not use that time point. The chapter also discusses the difference between *wide format* and *long format* of data. Wide format is often used for repeated measures ANOVA while long is used for MLM modeling. The long format gives more flexibility for varying occasions. 

An example of wide format:

```{r, echo=FALSE, comment="  ## "}
widedata = data.frame("ID"=c(1,2,3,4), 
                      "Outcome1"=c(32, 33, 40, 29),
                      "Outcome2"=c(33, 29, 42, 35),
                      "TimeVary1"=c(2,3,2,4),
                      "TimeVary2"=c(2,3,3,5),
                      "TimeInvariant"=c(0,1,1,0))
widedata
```

An example of long format:

```{r, echo=FALSE, comment="  ## "}
longdata = data.frame("ID"=c(1,1,1,2,2,2,2), 
                      "Outcome"=c(32, 33, 40, 29, 29, 42, 40),
                      "TimeVarying"=c(2,3,2,4,4,3,5),
                      "TimeInvariant"=c(0,0,0,1,1,1,1))
longdata
```

The chapter finishes with discussing analyzing patterns of change, issues with missing values, accelerated cohort designs, and metrics of time.

The typical form of longitudinal multilevel modeling is shown in the equation below:
$$
Y_{ti} = \beta_{00} + \beta_{10} Time_{ti} + \beta_{20} X_{ti} + \beta_{01} Z_i + u_{0i} + e_{ti}
$$
where $X_{ti}$ is a time-varying covariate (e.g., hours of sleep the night before) and $Z_i$ is a time-invariant covariate (e.g., race).


## Fixed vs. Varying Occasions

Whether it is fixed or varying occasions for measurements is not very important for MLM; it is extremely important for the ANOVA family of methods.

### Fixed Occasions

This design is where:

1. there are set time periods to measure each participant, 
2. each participant has the same number of measures, and 
3. the data is usually stored in wide format (although to use MLM, long format is generally required).

The ANOVA methods, Latent Growth Curve modeling [^latent] and MLM can be used here. When the design is balanced, both ANOVA and MLM methods will provide the same results. If there is no missing values, repeated measures ANOVA and MANOVA can be used to analyze this type of data. Repeated measures ANOVA can test the null that all time points are equal. It requires *sphericity*. A common form of sphericity is compound symmetry which requires that the variances are equal across time points and the covariances between time points are equal. 

Another ANOVA family method is MANOVA, which is much more complicated since it uses a transformation of the repeated measures and then does contrasts to analyze effects. It does not require sphericity, however. Unfortunately, time-varying covariates are not easily incorporated into the modeling.

However, MLM has distinct advantages here. Time-varying covariates can easily be added to the model, no sphericity assumptions are required, the intra-class correlation can be obtained from the intercept only model, and each person can have their own regression line/curve.

The section goes on to give an example of MLM on longitudinal data. It walks through the various model building, but focuses on interpreting the model, including the variance and covariance components (see page 89 and 90 for a discussion on interpreting the variance and the covariance simply). For example, the covariance (or if standardized, the correlation) tells us the relationship between the inital values and the growth. If there are covariates in the model in addition to the time variable, then this correlation is a partial correlation while controlling for the other predictors in the model.

### Varying Occasions

This design is where there are varying number of measures, at potentially different time points, and the data is best stored in long format. This section provides an example. Not surprisingly, it does not differ much from that of the previous section on fixed occasions [^fixed].


### Summary of Advantages of MLM

1. We can have growth curves that are fitted to each subject individually.
2. It can handle there being differing numbers of measurements for each individual.
3. Covariances between repeated measures can be modeled.
4. ANOVA is just a special case of MLM (with balanced data).
5. It's easy to add higher level effects (environmental or societal factors).
6. It's easy to incorporate time-varying covariates.

### Modeling Covariances

Can test different covariance patterns (from compound symmetry to unstructured). This is dicussed on page 105.

### Analyzing Patterns of Change

Several methods can be used in MLM to model non-linear relationships over time.

* Polynomial terms
  + With *k* fixed occasions, the polynomial can be to the *k-1* power
* Other shapes (logistic, exponential)
* Piecewise linear functions and splines

### Missing Values

MANOVA only measures complete cases (unless there was some form of imputation). MLM uses all the data points provided and, therefore, does not do case-wise exclusion. In this way, MANOVA assumes that missingness is completely at random (very restricted); whereas, MLM assumes missing at random (the missing can be related to other variables in the model). This is a very helpful feature, as missing at random is much less strict of an assumption.


### Accelerated Cohort Designs

Assuming the different cohorts in the sample are similar, we can combine the "time points" (e.g., ages) from the different subjects to get a more complete picture. For example, say person 1 was measured when he/she was 7, 9, and 11. Person 2 was measured when he/she was 10, 12, and 14. Using this design, we can model a curve from time point (age) 7 to time point 14. Instead of just the 4 years of the study, we can get a more complete picture of 7 years.


## Conclusion

That is all for my notes on Chapter 5. I recommend it for comparisons of MLM with other standard methods as well as examples on performing the analyses.


[^start]: These notes are for my Multilevel Modeling course in my doctoral program. This book by Hox is one I highly recommend. I thought I'd post it online in an easily accessible manner. If it helps you, let me know <t.barrett@aggiemail.usu.edu>. 

[^latent]: This is an Structural Equation Modeling technique that "models a repeated measures polynomial analysis of variance" (p. 81).

[^fixed]: This is not surprising since MLM doesn't really care if it is fixed or varying occasions.

