---
output: 
  pdf_document:
    latex_engine: xelatex
    includes: 
      in_header: preamble.tex
---

```{r scholar_info, echo = FALSE, warning = FALSE, message = FALSE}
library(scholar)
library(tidyverse)

your.id <- "UWdXXEoAAAAJ" # This is your google scholar ID
profile <- get_profile(your.id)
pubs <- get_publications(your.id, cstart = 0, pagesize = 100, flush = FALSE)
citations <- get_citation_history(your.id)

## Custom Code to make bib file from scholar ID
# url("https://scholar.googleusercontent.com/citations?view_op=export_citations&user=UWdXXEoAAAAJ&citsig=AMD79ooAAAAAXJvG8fCoULONM8SZwt9CLrw1g3CGG9nm&hl=en") %>% 
#   readLines() %>% 
#   paste("\n") %>% 
#   cat(file = "pubs.bib")

today <- lubridate::today()
month <- months(today)
day <- lubridate::day(today)
year <- lubridate::year(today)
```


<!-- Name -->
\hspace{.08in}\huge\textsc{Tyson S. Barrett, PhD} \hfill \footnotesize Updated: `r month` `r day`, `r year`
\normalsize
\vspace{8pt}

<!-- General Info -->
\begin{tabular}{rl}
    \textsc{Address:}     & 2800 Old Main Hill, Logan, UT 84322 \\
    \textsc{Email:}       & \href{mailto:tyson.barrett@usu.edu}{tyson.barrett@usu.edu} \\
    \textsc{Website:}     & \href{http://tysonbarrett.com}{tysonbarrett.com} \\
    \textsc{ORCID:}       & \href{https://orcid.org/0000-0002-2137-1391}{0000-0002-2137-1391} \\
    \textsc{OSF:}         & \href{osf.io/gkjbx}{osf.io/gkjbx} \\
    \textsc{Citations:}   & `r profile$total_cites` \\
    \textsc{H Index:}     & `r profile$h_index`
\end{tabular}

<!-- Education -->
\vspace{0pt}
\section{EDUCATION}
<!--\vspace{-5pt}-->
\hrule
\vspace{2pt}

\begin{tabular}{l|p{8cm}}
 \textsc{\textbf{Ph.D. Quantitative Psychology }} & \emph{Feb 2018}\\
 & \emph{Utah State University}\\
 &\footnotesize{\emph{\href{https://www.researchgate.net/publication/323946090_Marginal_Mediation_Analysis_A_New_Framework_for_Interpretable_Mediated_Effects?channel=doi&linkId=5ab418a60f7e9b4897c79cab&showFulltext=true}{Dissertation: Marginal Mediation Analysis: A New Framework for Interpretable Mediated Effects}}}\\
 \multicolumn{2}{c}{} \\
 \multicolumn{2}{c}{} \\

 \textsc{\textbf{B.S. Economics}} & \emph{May 2014} \\
  & \emph{Utah State University}\\
 &\footnotesize{\emph{Cum Laude}}\\
 \multicolumn{2}{c}{} \\
\textsc{\textbf{B.S. Psychology}} & \emph{May 2014} \\
 & \emph{Utah State University}\\
&\footnotesize{\emph{Cum Laude}}
\end{tabular}

<!-- Research Areas -->
\vspace{1pt}
\section{RESEARCH AREAS}
<!--\vspace{-5pt}-->
\hrule
\vspace{2pt}

\indent \hspace*{.1in} \small\textsc{Primary:} Accessible Methodology Using R Programming, Applied Methodology in the Disability Disciplines

\indent \hspace*{.1in} \textsc{Secondary:} Geographic Information Systems and Large-Scale Human Data

\normalsize

<!-- Research Experience -->
\vspace{1pt}
\section{RESEARCH APPOINTMENTS}
<!--\vspace{-5pt}-->
\hrule
\vspace{2pt}

\indent \hspace*{.1in} \small\textbf{Research Assistant Professor}, Department of Psychology, Utah State University \hfill 2018-Present
\newline \hspace*{.6in} \emph{Principle Investigator of the Social Data Lab}
\newline \hspace*{.6in} \emph{Managing Director of the Data Science \& Discovery Unit}

\indent \hspace*{.1in} \small\textbf{Research Assistant}, Prevention Science Lab, Utah State University \hfill 2016-2018
\newline \hspace*{.6in} \emph{Advisor: Ginger Lockhart, PhD}

\indent \hspace*{.1in} \small\textbf{Statistical and Data Science Consultant}, Utah State University \hfill 2016-2018
\newline \hspace*{.6in} \emph{Supervisors: Sarah Schwartz, PhD and Jamison Fargo, PhD}

\indent \hspace*{.1in} \small\textbf{WOC Research Assistant}, Veteran Affairs SLC \hfill 2016-2018
\newline \hspace*{.6in} \emph{Advisors: Vanessa Stevens, PhD and Richard Nelson, PhD}

<!-- Affiliations -->
\vspace{1pt}
\section{PROFESSIONAL SERVICE}
<!--\vspace{-5pt}-->
\hrule
\vspace{2pt}

\indent \hspace*{.1in} \small\textbf{\href{https://osf.io/gkjbx}{Center for Open Science }} Ambassador \hfill 2019-Present

\indent \hspace*{.1in} \small\textbf{\href{https://github.com/Rdatatable/data.table}{Rdatatable }} Contributor \hfill 2019-Present

\indent \hspace*{.1in} \small\textbf{\href{https://www.apa.org/pubs/journals/met}{Psychological Methods}} Voluntary Reviewer \hfill 2018-Present

\indent \hspace*{.1in} \small\textbf{\href{https://www.apa.org/pubs/journals/met}{Journal of Speech, Language, and Hearing Research}} Voluntary Reviewer \hfill 2018-Present


<!-- Grants -->
\vspace{10pt}
\noindent
\section{EXTERNAL FUNDING \normalsize (Ongoing)}
<!--\vspace{-5pt}-->
\hrule
\vspace{-2pt}

\setlength{\parindent}{-.3in}
\setlength{\leftskip}{.5in}
\setlength{\parskip}{5pt}
\noindent

\normalsize

\emph{Necessary Education and eXperiential Training} \hfill September 1, 2020 - August 30, 2024 \newline \emph{for Upcoming Para-professionals (NEXT-UP)}

\indent \hspace{.25in} Voss, M. (PI) & \textbf{Barrett, T.S. (Co-I)}. Sponsored by HRSA, 1 T26HP39463-01-00. Current Status: Funded. $2,296,000.

\indent \hspace{.25in} \emph{This project enhances Utah's peer workforce by addressing the clinical priorities of: 1) incorporating family-focused curriculum enhancements and 2) emphasizing and funding hands-on learning in mentored inter-professional field placements. It will increase training by more than 50\%, with basic and advanced (family-focused) certification and work experience. The evaluation of the program will be ongoing throughout the program's implementation.}

\vspace{8pt}

\emph{Video- and App-Based Naturalistic Language Instruction} \hfill July 1, 2020 -	June 30, 2024 \newline
\emph{(VALI) for Spanish-Speaking Caregivers to Support \newline Bilingual Language Development in Children \newline With or At Risk for Language Delays} 

\indent \hspace{.25in} Larson, A. (PI), Snyder, P. (Mentor) Domenech Rodriguez, M. (Mentor), & \textbf{Barrett, T.S. (Consultant)}. Sponsored by IES/NCSER, R324B200009. Current Status: Funded. $499,927.

\indent \hspace{.25in} \emph{The purpose of the research plan is to adapt an existing app, Háblame Bebé, that provides information to Spanish-speaking families about the importance of frequent and responsive verbal interactions with children in the home language but does not provide information on how to embed NLIS within everyday activities and routines to support DLLs with language delay.}

\vspace{8pt}

\emph{Application of Ideal Binary Masking to Disordered Speech} \hfill April 1, 2020 -	March 31, 2023 

\indent \hspace{.25in} Leopold, S. (PI), Borrie, S.A. (Co-I), & \textbf{Barrett, T.S. (Co-I)}. Sponsored by NIH/NIDCD, 1R21DC018641-01. Current Status: Funded. $430,000.

\indent \hspace{.25in} \emph{The major goal of this project is to utilize an established signal processing technique to investigate its potential application to overcome speech-in-noise difficulties for listeners understanding dysarthric speech.}


\vspace{8pt}

\emph{Perceptual training for improved intelligibility of dysarthric speech} \hfill July 1, 2020 - June 30, 2022

\indent \hspace{.25in} Borrie, S.A. (PI), Lansford, K. (PI), & \textbf{Barrett, T.S. (Co-I)}. Sponsored by NIH/NIDCD	1R21DC018867-01. Current Status: Funded. $420,000.

\indent \hspace{.25in} \emph{The major goal of this project is the systematic analysis and identification of speaker and listener parameters, and their interactions, that allow perceptual training paradigms to be optimized for intelligibility outcomes in dysarthria rehabilitation.}


\vspace{8pt}

\emph{Psychological flexibility in adults with ASD:} \hfill Sept 1, 2020 - Aug 31, 2021 \newline
\emph{Psychological distress in the context of \newline COVID-19} 

\indent \hspace{.25in} Aller, T.B. (PI), \textbf{Barrett, T.S. (Co-I)}, McClain Verdoes, M. (Co-I), \& Levin, M. (Co-I). Sponsored by National Research Consortium on Mental Health in Intellectual and Developmental Disabilities. Current Status: Funded. $10,000.

\indent \hspace{.25in} \emph{The aims of the study are to explore how psychological flexibility influences psychological distress in adults with ASD and to establish the dimensionality, internal consistency, test-retest reliability, and the concurrent and convergent validity of the Acceptance and Action Questionnaire II (AAQ-II) and of the Brief Experiential Avoidance Questionnaire (BEAQ) in adults with ASD.}


<!-- Published Works -->
\vspace{10pt}
\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
\setlength{\parskip}{0pt}
\noindent
\section{PEER-REVIEWED PUBLISHED WORKS}
\vspace{-5pt}
\hrule
\vspace{3pt}

\normalsize

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(stringr)
library(dplyr)
pubs = rio::import("pubs.xlsx", sheet = "Journals")
pubs_list = function(data){
  data = data %>% 
    mutate(across(everything(), ~furniture::washer(.x, is.na, value = ""))) %>% 
    group_by(year) %>% 
    mutate(header = if_else(row_number() == 1, as.character(year), " ")) %>% 
    mutate(
      authors = str_replace(authors, "Barrett, T.S.", "\\\\textbf{Barrett, T.S.}"),
      authors = str_replace(authors, "&", "\\\\&"),
      journal = paste0("\\emph{", journal, "}"),
      journal = if_else(volume == "", journal, paste0(journal, ", ")),
      year = paste0("(", year, ")."),
      journal_volume = if_else(volume == "", journal, paste0(journal, "\\emph{", volume, "}")),
      journal_volume_issue = if_else(issue != "", paste0(journal_volume, "(", issue, ")"), journal_volume),
      journal_volume_issue_pages = if_else(pages != "", 
                                           paste0(journal_volume_issue, ", ", pages, "."), 
                                           paste0(journal_volume_issue, "."))
    ) %>% 
    mutate(doi = str_replace(doi, "_", "\\_")) %>% 
    ungroup()
  
  cat("\\begin{longtable}{r!{\\color{lightdivider}\\vrule}p{14.6cm}}", "\n")
  rows = nrow(data)
  for (i in 1:nrow(data)){
      with(
        data %>% filter(row_number() == i),
          cat(
            header, " & \\hangindent=3em ",
            authors,
            tolower(year),
            title,
            journal_volume_issue_pages,
            opensci,
            if_else(i == rows, "\\vspace{5pt} \n", "\\vspace{5pt} \\\\ \n"),
            sep = " "
          )
      )
  }
  cat("\\end{longtable}")
}
```

```{r, echo=FALSE, results='asis'}
pubs_list(pubs)
```



<!-- WestEd Publications -->
\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
\setlength{\parskip}{0pt}
\noindent


\section{POLICY REPORTS}
\vspace{-5pt}
\hrule

\setlength{\parindent}{-.3in}
\setlength{\leftskip}{.5in}
\setlength{\parskip}{5pt}
\noindent

\normalsize

Willis, J., Doutre, S.M., Krausen, K., \textbf{Barrett, T.}, Ripma, T., & Caparas, R. (2020). \emph{California Special Education Funding System Study: A Descriptive Analysis of Special Education Funding in California}. WestEd. https://www.wested.org/resources/ca-special-education-funding-system/


<!-- Pre-Prints -->
\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
\setlength{\parskip}{0pt}
\noindent

\vspace{10pt}
\section{PUBLISHED PRE-PRINTS}
\vspace{-5pt}
\hrule
\vspace{3pt}
\footnotesize \hspace{4pt} Pre-prints not otherwise published in a peer-reviewed journal \normalsize

\setlength{\parindent}{-.3in}
\setlength{\leftskip}{.5in}
\setlength{\parskip}{5pt}
\noindent

\textbf{Barrett, T.S.} (2019, Oct 3). List-columns in `data.table`: Nesting and unnesting data tables and vectors. \emph{PsyArXiv}. https://doi.org/10.31234/osf.io/u8ekc \img{osfdata.png}{https://osf.io/smcg8/}\img{osfmaterials.png}{https://osf.io/smcg8/}

\textbf{Barrett, T.S.} (2019, August 15). Six Reasons to Consider Using R in Psychological Research. \emph{PsyArXiv}. https://doi.org/10.31234/osf.io/8mb6d \img{osfdata.png}{https://osf.io/436d7/}\img{osfmaterials.png}{https://osf.io/436d7/}

\textbf{Barrett, T.S.}, Cruz, R.A., \& Lockhart, G. (2019, June 18). Marginal Mediation Analysis: A Practical Statistical Framework for Interpretable Mediated Effects. \emph{PsyArXiv}. https://doi.org/10.31234/osf.io/fgm8t \img{osfdata.png}{https://osf.io/753kc/}\img{osfmaterials.png}{https://osf.io/753kc/}

\textbf{Barrett, T.S.}, Borrie, S.A., & Yoho, S.E. (2019, June 14). Automating with Autoscore: Introducing an R package for automating the scoring of orthographic transcripts. \emph{PsyArXiv}. https://doi.org/10.31234/osf.io/d8au4 \img{osfdata.png}{https://osf.io/htqvr/}\img{osfmaterials.png}{https://osf.io/htqvr/}


<!-- Published Works in Conference Proceedings -->
\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
\setlength{\parskip}{0pt}
\noindent

\vspace{10pt}
\section{PEER-REVIEWED CONFERENCE PROCEEDINGS}
\vspace{-5pt}
\hrule

\setlength{\parindent}{-.3in}
\setlength{\leftskip}{.5in}
\setlength{\parskip}{5pt}
\noindent

\normalsize

Lubold, N. Borrie, S.A., \textbf{Barrett, T.S.}, Willi, M., & Berisha, V. (2019). Do Conversational Partners Entrain on Articulatory Precision? \emph{Proceedings of INTERSPEECH 2019.} Paper number 1786, 1–5. (three reviewers, 50\% acceptance rate). doi: 10.21437/Interspeech.2019-1786

Willi, M., Borrie, S.A., \textbf{Barrett, T.S.}, Tu, M. & Berisha, V. (2018). A discriminative acoustic-prosodic approach for measuring local entrainment. \emph{Proceedings of INTERSPEECH 2018.} Paper number 1419, 1–5. (four reviewers, 50\% acceptance rate).

\vspace{-4pt}
\forceindent $-$ ArXiv Pre-Print: [1804.08663](http://arxiv.org/abs/1804.08663)


\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
\setlength{\parskip}{0pt}
\noindent


<!-- Other Publications -->
\vspace{20pt}
\section{OTHER PUBLISHED WORKS}
\vspace{-5pt}
\hrule

\setlength{\parindent}{0in}
\setlength{\leftskip}{.25in}
\setlength{\parskip}{0pt}
\noindent
\subsection{R Packages}
\setlength{\parindent}{-.5in}
\setlength{\leftskip}{1in}
\noindent

\textbf{Barrett, T.S.} (2019). `dissertateUSU`. URL: github.com/tysonstanley/dissertateUSU. R package version `r packageVersion("dissertateUSU")`. doi: 10.5281/zenodo.3240640. \img{dissertateUSU_hex.jpg}{https://github.com/TysonStanley/dissertateUSU}

\textbf{Barrett, T.S.} (2019). `tidyfast`. URL: github.com/tysonstanley/tidyfast. R package version `r packageVersion("tidyfast")`. \img{tidyfast_hex.png}{https://github.com/TysonStanley/tidyfast}

\textbf{Barrett, T.S.} (2019). `connectivity`. URL: github.com/tysonstanley/connectivity. R package version `r packageVersion("connectivity")`. \img{connectivity_hex.jpg}{https://github.com/TysonStanley/connectivity}

\textbf{Barrett, T.S.} (2019). `rjasa`. URL: github.com/tysonstanley/rjasa. R package version `r packageVersion("rjasa")`.

\textbf{Barrett, T.S.}, Borrie, S.A., & Yoho, S.E. (2019). `autoscore`. URL: github.com/autoscore/autoscore. R package version `r packageVersion("autoscore")`. \img{autoscore_hex.jpg}{https://github.com/autoscore/autoscore}

\textbf{Barrett, T.S.} (2018). `MarginalMediation`: Marginal Mediation Analysis. URL: tysonbarrett.com/MarginalMediation. R package version `r packageVersion("MarginalMediation")`. \img{mma_hex.jpg}{https://github.com/TysonStanley/MarginalMediation}

\textbf{Barrett, T.S.}, Brignone., E. \& Laxman, D. (2018) `furniture`: Furniture for Quantitative Researchers. URL: tysonbarrett.com/furniture. R package version `r packageVersion("furniture")`. \img{furniture_hex.jpg}{https://github.com/TysonStanley/furniture}


\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
\setlength{\parskip}{0pt}

\vspace{10pt}
\section{CITATION COUNT HISTORY}
\vspace{-5pt}
\hrule
\vspace{20pt}

```{r citation_fig, fig.width=8, fig.height=2.3, echo = FALSE, warning = FALSE, message = FALSE}
month <- months(lubridate::today())
year <- lubridate::year(lubridate::today())

ggplot(citations, aes(year, cites, fill = year)) +
  geom_col(color = "black") +
  geom_text(aes(label = cites),
            nudge_y = 10) +
  labs(x = "Year", 
       y = "Citation Count\n(by Year)",
       caption = paste("Note: Data for citations in", year, "until", month)) +
  theme_classic() +
  scale_fill_viridis_c(direction = -1,
                        begin = .1, 
                        end = .9) +
  theme(legend.position = "none",
        axis.line = element_line(color = "black", size = .4))
```




<!-- Selected Presentations -->
\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
\setlength{\parskip}{0pt}
\noindent
\section{SELECTED PRESENTATIONS}
\vspace{-5pt}
\hrule
\setlength{\parindent}{-.3in}
\setlength{\leftskip}{.5in}
\setlength{\parskip}{6pt}
\noindent

<!-- -->

Smith, B.M., Barney, J.L., Ong, C.W., \textbf{Barrett, T.S.}, Levin, M.E., & Twohig, M.P. (2020, November). Outcomes and mechanisms of change in acceptance and regulation approaches to exposures for intrusive thoughts. In S. P. Buinewics & J. Buchholz (Chairs), Understanding the therapeutic process of exposure therapy to optimize treatment. Paper presented at the annual meeting of the Association for Behavioral and Cognitive Therapies, online.

Smith, B. M., Barney, J.L., Ong, C.W., \textbf{Barrett, T.S.}, Levin, M.E., & Twohig, M.P. (2020, July). Physiological, behavioral, and self-report outcomes of acceptance and regulation approaches to exposures for intrusive thoughts. In S. Mpoulkoura (Chair), Third-wave approaches to anxiety disorders treatment: Outcomes and processes of change. Paper presented at the annual meeting of the Association for Contextual Behavioral Science, online.

Hancock, A.S., Orellana, C., Alphonsa, S., \textbf{Barrett, T.S.}, Logan, K., Gillam, S.L., & Gillam, R.B. (2020, May). An fNIRS Investigation During Stuttered and Fluent Speech. Poster presentation at Cognitive Neuroscience Society, Boston, Massachusetts.

**Barrett, T.S.** (2020, January). Working with list-columns in `data.table`. Talk presented at the RStudio Conference 2020, San Francisco, CA.

Vázquez, A.L., Chou, T., Navarro, C., **Barrett, T.S.**, Villodas, M.T., & Domenech Rodríguez, M.M. (2019, November). Understanding caregiver perceived need for and utilization of adolescent counseling services through machine learning. In M. D. Hetzel-Riggin (Chair). Addressing violence, abuse, and trauma throughout the socioecological model. Symposium accepted for presentation at the Association for Behavioral and Cognitive Therapies, Atlanta, GA.

Budge, S., Borrie, S.A., **Barrett, T.S.**, Yoho, S.E., Vargas, R., & Lansford, K. (2019, November). Gamified Perceptual Learning of Dysarthric Speech. Poster presented at the 2019 ASHA Convention, Orlando, FL.

Hirsch, M., Lansford, K., Borrie, S.A., & **Barrett, T.S.**. (2019, November). Familiarization of Dysarthric Speech: Generalization Between Speakers of Different Sex. Poster presented at the 2019 ASHA Convention, Orlando, FL.

Ong, C.W., Lee, E.B., Krafft, J., Terry, C.L., **Barrett, T.S.**, Levin, M.E., & Twohig, M.P. (2019, November). A randomized controlled trial of acceptance and commitment therapy for clinical perfectionism. In C. W. Ong (Chair), Using acceptance and commitment therapy for different clinical presentations: Evidence-based adaptations. Symposium conducted at the 53rd annual convention of the Association for Behavioral and Cognitive Therapies, Atlanta, GA.
 
Vázquez, A.L., **Barrett, T.S.**, Domenech Rodríguez, M.M., Bustos, M., Gutierrez, M., Amador Buenabad, N.G., & Villatoro Velazquez, J.A. (2019, March). The influence of perceived parenting on substance initiation among Latinx children. Poster presented at the Society for Research in Child Development, Baltimore, MD.
 
Vázquez, A.L., Chou, T., Navarro, C., **Barrett, T.S.**, & Domenech Rodríguez, M.M. (2019, March). Examining caregiver perceived need for and utilization of adolescent counseling services through machine learning. Poster presented at the Society for Research in Child Development, Baltimore, MD.

Vázquez, A. L., Domenech Rodríguez, M. M., **Barrett, T.S.**, Amador Buenabad, N., & Villatoro Velazquez, J. A. (2018, September) Factores de crianza que tienen alto impacto sobre el consumo de drogas por parte niños. In M. M. Domenech Rodríguez (Chair). Crianza positiva y consumo de drogas: hallazgos, modelos y retos para la implementación. Symposium presented at the 2o Congreso Mundial de Prevención de Adicciones en Niños y Adolescentes, Tijuana, BCS, MX.

**Barrett, T.S.**, & Lockhart, G. (2017). Enhancing the Exploration and Communication of Big Data in Prevention Science. Poster presented at the Annual Meeting of the Society of Prevention Research, Washington, DC. *Received "Distinguished Poster Award" and "Abstract of Distinction."*

**Barrett, T.S.**, & Lockhart, G. (2017). Exploring the Predictors of Marijuana Use Among Adolescents with Asthma. Oral presentation at the Utah State University Research Symposium, Logan, UT.

Sanghavi, K., White, K., **Barrett, T.S.**, Wylie, A., Raspa, M., Cashman, D., Vogel, B. Caggana, M. & Bodurtha, J. (2017). Poster presented at the Early Hearing Detection and Intervention Conference, Atlanta, GA. *Received "Outstanding Poster Award."*

Brignone, E., Gundlapalli, A.V., **Barrett, T.S.**, Blais, R.K., Nelson, R.E., Carter, M.E., Kimerling, R., Samore, M.H., Fargo, J.D. (2016). Cost of Care among Male and Female Veterans with a Positive Screen for Military Sexual Trauma. Poster presented at the 2016 Annual Meeting of the International Conference of Psychology, Yokohama, Japan.

**Barrett, T.S.**, Munoz, K. & White, K. (2016). How well do parent report hearing loss in their children? Poster presented at the Early Hearing Detection and Intervention Conference, San Diego, CA.

**Barrett, T.S.**, Munoz, K. & White, K. (2016). Accounting for Temporary Loss in National Studies on Hearing Loss. Poster presented at the Early Hearing Detection and Intervention Conference, San Diego, CA.

**Barrett, T.S.**, Munoz, K. & White, K. (2016). An Evaluation of Early Intervention delivered via Video Conferencing. Poster presented at the Early Hearing Detection and Intervention Conference, San Diego, CA.

Stevens, V., **Barrett, T.S.** & Nelson, R. (2016). Distribution and Daily Cost of Care in a Pediatric Hospital. Oral presentation to the Pediatric Guidance Council of Intermountain Healthcare, Salt Lake City, UT.

**Barrett, T.S.**, Munoz, K. & White, K. (2015). Refinements to estimating prevalence of hearing loss in children. Poster presented at the Utah State University Research Symposium, Logan, UT.

**Barrett, T.S.**, Prante, M., Peterson, R., Fargo, J.D., Pyle, N. (2014). Predictors of employability among homeless youth. Poster presented at the Psi-Chi Undergraduate Research Conference at Idaho State University, Pocatello, ID. *Best Undergraduate Poster Presentation Award.*

**Barrett, T.S.**, Holland, D. (2014). Nascent Entrepreneurship, Impulsivity, and Self- Efficacy. Poster presented at the Research on Capitol Hill, Salt Lake City, UT.

Holland, D., **Barrett, T.S.** (2013). Impulsivity in young entrepreneurs. Round table discussion at the Babson Business Conference, Paris, France.

\vspace{10pt}
\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
\setlength{\parskip}{0pt}
\noindent
\subsection{\textsc{Professional Development Activities Attended}}
\vspace{-5pt}
\hrule
\setlength{\parindent}{-.5in}
\setlength{\leftskip}{1in}
\setlength{\parskip}{6pt}
\noindent

RStudio Conference sponsored by RStudio. Training on advanced features of R and RStudio, including training on data wrangling, reproducible analyses, and database access. January 2019.

RStudio Conference sponsored by RStudio. Special training on using the Shiny developer interface for creating interactive, responsive websites using R. January 2020.

American Educational Research Association Annual Meeting (Virtual). Training on quantitative methodologies (including psychometrics and econometrics) for studying educational policy and practice. April 2021.


<!-- Awards -->
\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
\setlength{\parskip}{0pt}
\noindent
\vspace{10pt}
\section{AWARDS}
\vspace{-5pt}
\hrule
\vspace{10pt}

- Abstract of Distinction (Annual Meeting of the Society of Prevention Research)
- Distinguished Poster Award (Annual Meeting of the Society of Prevention Research)
- Outstanding Poster Award (Annual Meeting of EHDI)
- Dean’s Scholarship (full tuition for two years)
- Best Poster Presentation (Psi-Chi Undergraduate Conference)


<!-- Teaching Interests -->
\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
\setlength{\parskip}{0pt}
\noindent
\vspace{10pt}
\section{TEACHING INTERESTS}
\vspace{-5pt}
\hrule
\vspace{10pt}

\subsection{\textsc{Primary Interest: Quantitative Methods}}

- Undergraduate and Graduate Applied Statistics
- Multilevel Modeling (Hierarchical Linear Modeling, GEE, Mixed Effects)
- Reproducible Research (Research Methods, Open Science Framework)
- R for the Social Sciences (Undergraduate and Graduate Level)
- Research Methods (Undergraduate and Graduate Level)
- Causality and Econometrics (Undergraduate and Graduate Level)
- Structural Equation Modeling (Psychometrics and Measurement Models, Mixture Modeling)
- Mediation Analysis (Marginal Mediation, Moderated Mediation)


<!-- Teaching Experience -->
\section{TEACHING EXPERIENCE}
\vspace{-5pt}
\hrule
\vspace{10pt}

- **Applied Statistical Analysis** (EDUC 6050)
    - Graduate-level applied statistics course for non-research degrees

- **Research Design \& Analysis I** (EDUC 6600)
    - Graduate-level applied statistics course for PhD students

- **Research Design \& Analysis II** (EDUC 7610)
    - Graduate-level linear regression course for PhD students

- **R for the Health, Behavioral, Educational, and Social Sciences I and II** (EDUC 6560)
    - Created, Developed, and Taught
    - Graduate Level

- **Research Methods** (PSY 3500)
    - Undergraduate-level psychological research methods
        - Online
        - In-person


\subsubsection{\hspace{10pt} \emph{MOST RECENT TEACHING EVALUATIONS (Fall 2020)}}

\begin{center}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
       & Much  &       &         &        & Much   \\
Course & Lower & Lower & Similar & Higher & Higher \\
\hline
Research Methods (PSY 3500)                & & & & X & \\
Applied Statistical Analysis (EDUC 6050)   & & & & X & \\
R Course I \& II (Continuing Ed Credit)    & & & & X & \\
Statistical Foundations (EDUC 6600)        & & & & X & \\
Regression Analysis (EDUC 7610)            & & & & X & \\
\hline
\end{tabular}
\end{center}


<!-- Software and Programming Experience -->
\vspace{20pt}
\section{SOFTWARE AND PROGRAMMING EXPERIENCE}
\vspace{-5pt}
\hrule
\vspace{10pt}

1. **R and RMarkdown: Expert (9 years of daily use)**
    - Database Queries, Creation, and Management
    - Data Analytics
    - Website Creation
    - Program Development and Deployment 
2. **REDCap: Moderate Experience (4 year of weekly use)**
    - Database Creation
    - Data Entry and Survey Creation and Deployment
    - Data Management via REDCap API
3. **SQL: Some Experience (7 years of occasional use)**
    - Database Queries and Database Management 
    - Data Analytics
4. **Python: Minor Experience (1 year of occasional use)**
    - Data Analytics
    - MTurk customization
    
    


