---
title: "Difference-in-Differences"
subtitle: "From Classical Foundations to Modern Methods"
author: "DiD Workshop"
format:
  revealjs:
    theme: [default]
    incremental: false
    slide-number: true
    chalkboard: true
    preview-links: auto
    code-fold: false
    code-tools: true
    code-line-numbers: false
    fig-width: 10
    fig-height: 6
    scrollable: true
execute:
  echo: true
  warning: false
  message: false
  cache: false
---

```{r setup, include=FALSE}
# Load required packages
library(tidyverse)
library(fixest)
library(knitr)
library(kableExtra)

# Set theme for plots
theme_set(theme_minimal(base_size = 14))

# Set seed for reproducibility
set.seed(2025)
```

## Today's Overview {.smaller}

::: {.columns}
::: {.column width="25%"}
### Part 1: Foundations {.fragment}

- Potential outcomes framework
- Parallel trends assumption
- Classical 2×2 DiD design
:::

::: {.column width="25%"}
### Part 2: When DiD Fails {.fragment}

- Common pitfalls
- TWFE with staggered adoption
- Treatment effect heterogeneity
:::

::: {.column width="25%"}
### Part 3: Solutions {.fragment}

- Heterogeneity-robust methods
- Modern approaches (CS, SA, BJS)
:::

::: {.column width="25%"}
### Part 4: Implementation {.fragment}

- R examples
- Diagnostics & workflow
- Best practices
:::
:::
:::

# Part 1: Foundations {background-color="#2C5F8D"}

## What is Difference-in-Differences?

::: {.callout-note icon=false}
## Core Idea
Compare **changes over time** between treated and control groups
:::

::: {.columns}
::: {.column width="50%"}
**Treatment Group**

- Experiences the intervention
- Δ = After - Before
:::

::: {.column width="50%"}
**Control Group**

- Does not experience intervention
- Δ = After - Before
:::
:::

::: {.fragment .callout-tip icon=false}
## Treatment Effect
**(Treated After - Treated Before) − (Control After - Control Before)**

The "difference-in-differences" removes time-invariant confounding
:::

## Potential Outcomes Framework

::: {.callout-warning icon=false}
## The Fundamental Problem
We can only observe **one** potential outcome for each unit
:::

::: {.columns}
::: {.column width="50%"}
### $Y^1_{it}$
Outcome if unit $i$ is **treated** at time $t$
:::

::: {.column width="50%"}
### $Y^0_{it}$
Outcome if unit $i$ is **NOT treated** at time $t$
:::
:::

::: {.fragment}
### Individual Treatment Effect

$$\tau_{it} = Y^1_{it} - Y^0_{it}$$

**The Problem:** We never observe both $Y^1_{it}$ and $Y^0_{it}$ for the same unit at the same time

**DiD's Solution:** Use control group trends to estimate the missing counterfactual
:::

## The Parallel Trends Assumption

::: {.callout-important icon=false}
## Core Identifying Assumption
**Parallel Trends:** In the absence of treatment, the treatment and control groups would have followed parallel outcome trajectories
:::

$$E[Y^0_{i2} - Y^0_{i1} \mid D=1] = E[Y^0_{i2} - Y^0_{i1} \mid D=0]$$

::: {.columns}
::: {.column width="50%"}
### ✓ What It Allows

- Different baseline levels
- Time-invariant differences
- Common time shocks
:::

::: {.column width="50%"}
### ✗ What It Rules Out

- Differential time trends
- Group-specific shocks
- Pre-existing trend differences
:::
:::

::: {.fragment .callout-note}
This assumption is **untestable** but can be made more plausible through design and diagnostics
:::

## Example 1: Classic 2×2 DiD Design

**Scenario:** State implements minimum wage increase in 2010

```{r classic-did-data}
# Generate classic 2x2 DiD data
set.seed(123)
data_2x2 <- expand.grid(state = 1:10, year = 2009:2010)

data_2x2 <- data_2x2 %>%
  mutate(
    treat_state = ifelse(state <= 5, 1, 0),
    post = ifelse(year == 2010, 1, 0),
    treated = treat_state * post,
    # True effect = 5
    outcome = 50 + 10*treat_state + 2*post + 5*treated + rnorm(n(), 0, 2)
  )

# Summary statistics
data_2x2 %>%
  group_by(treat_state, post) %>%
  summarise(mean_outcome = mean(outcome), .groups = "drop") %>%
  pivot_wider(names_from = post, values_from = mean_outcome, 
              names_prefix = "period_") %>%
  mutate(delta = period_1 - period_0) %>%
  kable(digits = 2, col.names = c("Group", "Pre", "Post", "Δ"))
```

## Example 1: Estimation

```{r classic-did-estimate}
# Estimate DiD with TWFE
did_model <- feols(outcome ~ treated | state + year, 
                   data = data_2x2, cluster = ~state)

# Display results
etable(did_model, dict = c(treated = "Treatment Effect"))
```

::: {.callout-tip}
**DiD Estimate:** `r round(coef(did_model)[1], 2)` (true effect = 5)
:::

## Example 1: Visualization

```{r classic-did-viz, fig.width=10, fig.height=5}
#| code-fold: true
data_2x2 %>%
  group_by(treat_state, year) %>%
  summarise(mean_outcome = mean(outcome), .groups = "drop") %>%
  ggplot(aes(x = year, y = mean_outcome, 
             color = factor(treat_state, labels = c("Control", "Treated")),
             group = treat_state)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_vline(xintercept = 2009.5, linetype = "dashed", alpha = 0.5) +
  annotate("text", x = 2009.5, y = max(data_2x2$outcome), 
           label = "Treatment", vjust = -0.5, hjust = -0.1) +
  labs(title = "Classic 2×2 DiD: Parallel Trends",
       x = "Year", y = "Outcome", color = "Group") +
  scale_color_manual(values = c("Control" = "#E69F00", "Treated" = "#56B4E9")) +
  theme(legend.position = "bottom")
```

## Two-Way Fixed Effects (TWFE) Regression

::: {.callout-note icon=false}
## The Classic Approach
$$Y_{it} = \alpha_i + \phi_t + \beta \cdot D_{it} + \varepsilon_{it}$$
:::

::: {.columns}
::: {.column width="33%"}
### $\alpha_i$
**Unit fixed effects**

Controls for time-invariant differences between units
:::

::: {.column width="33%"}
### $\phi_t$
**Time fixed effects**

Controls for common time shocks affecting all units
:::

::: {.column width="33%"}
### $\beta$
**DiD estimate**

Our treatment effect of interest
:::
:::

::: {.fragment .callout-important}
## Why TWFE Works in 2×2 Designs

- Only one treatment timing (everyone treated at once or never)
- Single before/after comparison
- Treatment effect homogeneity not required
- Clean identification of average treatment effect
:::

# Part 2: When DiD Fails {background-color="#8B0000"}

## Common Pitfalls and How to Avoid Them {.smaller}

::: {.columns}
::: {.column width="50%"}
### ❌ What NOT to Do

- Using TWFE with staggered adoption
- Ignoring treatment effect heterogeneity
- Not checking pre-trends
- Wrong standard errors
- Including post-treatment as controls
- Testing multiple specs, reporting one
:::

::: {.column width="50%"}
### ✓ Best Practices

- Use heterogeneity-robust methods
- Always visualize pre-trends
- Report multiple aggregations
- Cluster SEs correctly
- Conduct sensitivity analysis
- Pre-register when possible
:::
:::

::: {.fragment .callout-warning}
### Warning Signs Your DiD May Be Invalid

- Pre-treatment effects are significant and large
- Effects change dramatically with specification changes
- Bacon decomposition shows heavy negative weights
- TWFE and robust methods give very different answers
:::

## But what happens with multiple treatment timings? {background-color="#8B0000"}

::: {.r-fit-text}
⚠️
:::

## The Problem: TWFE with Staggered Adoption

::: {.callout-warning icon=false}
## When units are treated at different times, TWFE can give **severely biased** estimates
:::

### Three Critical Problems {.smaller}

1. **Forbidden Comparisons:** Already-treated units used as controls
2. **Negative Weights:** Some 2×2 DiDs get negative weight  
3. **Wrong Target:** Doesn't estimate a meaningful average effect

::: {.fragment}
### Goodman-Bacon Decomposition

TWFE $\beta$ is a weighted average of:

- Earlier vs Never treated ✓
- Later vs Never treated ✓
- Earlier vs Not-yet treated ✓
- **Later vs Already treated** ✗ ← This causes problems!
:::

## Example 2: When TWFE Fails {.smaller}

**Scenario:** States adopt policy in different years with dynamic effects

```{r twfe-failure-data}
# Simulate staggered adoption with dynamic effects
set.seed(456)
data_staggered <- expand.grid(state = 1:50, year = 2010:2020)

data_staggered <- data_staggered %>%
  mutate(
    treat_year = case_when(
      state <= 20 ~ 2015,
      state <= 35 ~ 2018,
      TRUE ~ NA_real_
    ),
    time_to_treat = ifelse(!is.na(treat_year), year - treat_year, -1000)
  )

# True dynamic effects (growing over time)
data_staggered <- data_staggered %>%
  mutate(
    true_effect = case_when(
      time_to_treat < 0 ~ 0,
      time_to_treat == 0 ~ 3,
      time_to_treat == 1 ~ 8,
      time_to_treat == 2 ~ 12,
      time_to_treat >= 3 ~ 15,
      TRUE ~ 0
    ),
    outcome = 50 + 2*(year - 2010) + true_effect + rnorm(n(), 0, 5)
  )

# Show true effects
tibble(
  `Time Since Treatment` = c("Pre-treatment", "Year 0", "Year 1", "Year 2", "Year 3+"),
  `True Effect` = c(0, 3, 8, 12, 15)
) %>% kable()
```

## Example 2: TWFE Estimation (WRONG!)

```{r twfe-wrong}
# Traditional TWFE approach (BIASED!)
data_staggered <- data_staggered %>%
  mutate(treated = !is.na(treat_year) & year >= treat_year)

twfe_wrong <- feols(
  outcome ~ treated | state + year, 
  data = data_staggered, cluster = ~state
)

etable(twfe_wrong, dict = c(treated = "Treatment Effect (TWFE)"))
```

::: {.callout-warning}
**Problem:** TWFE gives us **one number** but true effects range from 3 to 15!

The estimate is biased because it uses "forbidden comparisons" and negative weights.
:::

## Example 2: Event Study Shows the Problem

```{r event-study-wrong, fig.width=10, fig.height=5}
#| code-fold: true
# Event study with TWFE (also biased)
event_study_wrong <- feols(
  outcome ~ i(time_to_treat, ref = -1) | state + year,
  data = data_staggered %>% filter(time_to_treat >= -5 & time_to_treat <= 5),
  cluster = ~state
)

# Plot event study
iplot(event_study_wrong, 
      main = "TWFE Event Study (Biased!)",
      xlab = "Time to Treatment",
      ylab = "Estimated Effect") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_vline(xintercept = -0.5, linetype = "dashed", color = "blue", alpha = 0.5)
```

::: {.callout-warning .smaller}
Notice: Effects don't match true values (0, 3, 8, 12, 15) due to negative weighting!
:::

## Why Treatment Effect Heterogeneity Matters

::: {.callout-important icon=false}
## Key Insight
When treatment effects vary across groups or over time, TWFE weights can be **negative**, causing severe bias
:::

::: {.columns}
::: {.column width="50%"}
### Homogeneous Effects

Effect is same for everyone

- TWFE works fine
- Negative weights don't matter
- Get correct average effect
:::

::: {.column width="50%"}
### Heterogeneous Effects {.fragment}

Effects differ by group/time

- TWFE can be severely biased
- Negative weights cause problems
- **Can even reverse signs!**
:::
:::

::: {.fragment .callout-note}
### Common Sources of Heterogeneity

- Effects grow/fade over time
- Different effect sizes by region
- Treatment intensity varies
- Anticipation effects differ
:::

# Part 3: Modern Solutions {background-color="#2C8B5C"}

## Heterogeneity-Robust Methods {.smaller}

::: {.callout-tip icon=false}
## Three leading approaches that handle staggered adoption correctly
:::

### 1. Callaway & Sant'Anna (2021) {.fragment}
- Estimates group-time ATT(g,t) using clean 2×2 comparisons
- **Pros:** Most flexible, doubly robust, multiple aggregations
- **Cons:** Can be less efficient
- **R Package:** `did`

### 2. Sun & Abraham (2021) {.fragment}
- Interaction-weighted estimator with cohort-specific indicators
- **Pros:** Easy to implement, works well for event studies
- **R Package:** `fixest::sunab()`

### 3. Borusyak, Jaravel & Spiess (2024) {.fragment}
- Imputation-based approach on not-yet-treated observations
- **Pros:** Most efficient under strong assumptions
- **R Package:** `did2s`

::: {.callout-note}
### When to Use Which Method?

- **Default choice:** Callaway-Sant'Anna (most robust)
- **Event studies:** Sun-Abraham (easy in `fixest`)
- **Maximum efficiency:** Borusyak-Jaravel-Spiess (strong assumptions)
:::

## What We Know Now {.smaller}

::: {.incremental}
1. **The Problem:** TWFE fails with staggered adoption + heterogeneous effects
   - Uses forbidden comparisons
   - Applies negative weights
   - Can severely bias estimates

2. **The Solution:** Modern heterogeneity-robust methods
   - Use only clean 2×2 comparisons
   - Properly handle treatment timing variation
   - Allow for treatment effect heterogeneity

3. **What To Do:** Use Callaway-Sant'Anna (or Sun-Abraham/Borusyak-Jaravel-Spiess)
   - Check pre-trends visually
   - Report multiple aggregations (overall, dynamic, by-group)
   - Conduct sensitivity analysis
:::

# Part 4: Implementation & Practice {background-color="#5B3A8F"}

## Example 3: Callaway-Sant'Anna Implementation {.smaller}

```{r cs-install, eval=FALSE}
# Install if needed
install.packages("did")
```

```{r cs-estimate}
library(did)

# Prepare data for CS method
data_cs <- data_staggered %>%
  filter(!is.na(treat_year) | state > 35) %>%
  mutate(treat_year = ifelse(is.na(treat_year), 0, treat_year))

# Estimate group-time effects
cs_results <- att_gt(
  yname = "outcome",
  tname = "year",
  idname = "state",
  gname = "treat_year",
  data = data_cs,
  control_group = "notyettreated",
  est_method = "dr"
)
```

## Example 3: CS Results - Overall ATT

```{r cs-overall}
# Overall ATT (simple average)
agg_simple <- aggte(cs_results, type = "simple")
summary(agg_simple)
```

::: {.callout-tip}
The overall ATT averages effects across all treated groups and time periods
:::

## Example 3: CS Results - Dynamic Effects

```{r cs-dynamic}
# Dynamic effects (event study)
agg_dynamic <- aggte(cs_results, type = "dynamic")
summary(agg_dynamic)
```

## Example 3: CS Event Study Visualization

```{r cs-event-plot, fig.width=10, fig.height=6}
#| code-fold: true
# Plot event study
ggdid(agg_dynamic, 
      xlab = "Time to Treatment",
      ylab = "ATT",
      title = "Callaway-Sant'Anna Event Study (Correct!)") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_minimal(base_size = 14)
```

::: {.callout-tip .smaller}
Now we see the true dynamic pattern: effects start at ~3 and grow to ~15!
:::

## Example 3: CS Results - By Group

```{r cs-group}
# Effects by treatment cohort
agg_group <- aggte(cs_results, type = "group")
summary(agg_group)
```

::: {.callout-note}
This reveals heterogeneity across treatment cohorts
:::

## Comparison: TWFE vs Callaway-Sant'Anna

```{r comparison-table}
# Create comparison
comparison <- tibble(
  Method = c("TWFE (Wrong)", "Callaway-Sant'Anna (Correct)", "True Average"),
  Estimate = c(
    coef(twfe_wrong)[1],
    agg_simple$overall.att,
    mean(data_staggered %>% 
           filter(!is.na(treat_year) & year >= treat_year) %>% 
           pull(true_effect))
  )
)

comparison %>%
  mutate(Estimate = round(Estimate, 2)) %>%
  kable()
```

::: {.callout-important}
TWFE is biased! Callaway-Sant'Anna recovers the true average effect.
:::

## Data Structure and Requirements

::: {.callout-note icon=false}
## Required Data Format: Panel Data
:::

### Required Variables

- **Unit ID:** Unique identifier for each unit (state, firm, individual)
- **Time variable:** Period indicator (year, quarter, month)
- **Outcome:** The dependent variable of interest
- **Treatment timing:** When each unit first receives treatment (NA for never-treated)

### Key Considerations

- **Treatment is absorbing:** Once treated, units stay treated (no switching off)
- **Never-treated units:** Coded with NA or 0 for `treat_year`
- **Pre-treatment periods:** Need sufficient periods before first treatment
- **Covariates:** Optional but can improve precision

## Diagnostic Testing: Pre-Trends

::: {.callout-warning icon=false}
## Parallel Trends Cannot Be Proven, But Can Be Made More Plausible
:::

### 1. Visual Inspection {.fragment}
Plot outcome trends for treated vs control groups before treatment

### 2. Formal Pre-Trend Tests {.fragment}
Test if pre-treatment effects are jointly zero

⚠️ **Warning:** These tests often have **low power**

### 3. Placebo Tests {.fragment}
- Use fake treatment dates in pre-period
- Test on untreated units only
- Should find no effect if parallel trends holds

## Diagnostic Testing: Sensitivity Analysis

```{r sensitivity-example, eval=FALSE}
# Sensitivity analysis with HonestDiD
# install.packages("HonestDiD")
library(HonestDiD)

# Create sensitivity analysis
# How large would trend violations need to be to overturn conclusions?
robust_ci <- createSensitivityResults(
  betahat = coef_vector,
  sigma = vcov_matrix,
  M = 0.5  # Bound on trend violations
)
```

::: {.callout-tip}
Instead of treating parallel trends as a binary assumption, ask:

**"How large would trend violations need to be to overturn my conclusions?"**
:::

## Standard Errors and Inference {.smaller}

::: {.callout-important icon=false}
## Standard errors are critical! Incorrect SEs can lead to invalid inference
:::

### Clustering Rules of Thumb

- **Default:** Cluster at the unit level (state, firm, individual)
- **Why?** Treatment typically assigned at unit level, creating within-unit correlation
- **Two-way clustering:** If treatment correlates across both units and time

```{r se-example, eval=FALSE}
# With fixest
library(fixest)
model <- feols(outcome ~ treatment | unit + time,
               data = data,
               cluster = ~ unit)  # Cluster at unit level

# With did package (automatic clustering)
cs_results <- att_gt(...)

# Bootstrap standard errors (alternative)
cs_boot <- att_gt(..., boot = TRUE, nboot = 1000)
```

## Recommended Analysis Workflow

1. **Prepare Your Data** → Panel format, treatment timing variable
2. **Visualize Treatment Patterns** → Plot timing, check variation
3. **Check Pre-Trends** → Event study plot, visual inspection
4. **Estimate with Robust Methods** → Use Callaway-Sant'Anna or Sun-Abraham
5. **Report Multiple Aggregations** → Overall ATT, dynamic effects, by-group
6. **Conduct Sensitivity Analysis** → Different control groups, HonestDiD
7. **Document Transparently** → All specs, robustness checks, limitations

::: {.callout-tip}
**When in doubt:** Use Callaway-Sant'Anna, check pre-trends, conduct sensitivity analysis
:::

## Essential R Packages for Modern DiD {.smaller}

::: {.columns}
::: {.column width="50%"}
### `did` ⭐
Callaway & Sant'Anna implementation
```r
install.packages("did")
```

### `fixest`
Fast fixed effects, Sun-Abraham
```r
install.packages("fixest")
```

### `bacondecomp`
Goodman-Bacon decomposition
```r
install.packages("bacondecomp")
```
:::

::: {.column width="50%"}
### `HonestDiD`
Rambachan & Roth sensitivity
```r
# From GitHub
devtools::install_github(
  "asheshrambachan/HonestDiD"
)
```

### `synthdid`
Synthetic DiD
```r
# From GitHub
devtools::install_github(
  "synth-inference/synthdid"
)
```
:::
:::

::: {.callout-tip}
💡 **Start with:** `did` for estimation, `fixest` for quick diagnostics
:::

# Key Takeaways {background-color="#2C5F8D"}

## The DiD Revolution Changed Everything

::: {.incremental}
1. **Traditional TWFE is often WRONG**
   - With staggered adoption and heterogeneous effects, TWFE can be severely biased

2. **Use heterogeneity-robust methods**
   - Callaway-Sant'Anna, Sun-Abraham, or Borusyak-Jaravel-Spiess are now best practice

3. **Parallel trends is CRITICAL**
   - Check pre-trends visually, conduct sensitivity analysis, be transparent

4. **Report heterogeneity**
   - Show dynamic effects, by-group effects, not just one number

5. **Standard errors matter**
   - Cluster at the treatment assignment level
:::

## {background-color="#2C8B5C"}

::: {.r-fit-text}
**Default choice:**

Use Callaway-Sant'Anna

with the `did` package
:::

## Additional Resources {.smaller}

### Key Papers
- **Goodman-Bacon (2021)** - "Difference-in-differences with variation in treatment timing"
- **Callaway & Sant'Anna (2021)** - "Difference-in-Differences with multiple time periods"
- **Sun & Abraham (2021)** - "Estimating dynamic treatment effects in event studies"
- **Roth et al. (2023)** - "What's trending in difference-in-differences"

### Books & Online Resources
- **The Effect** (Huntington-Klein) - Free online textbook
- **Causal Inference: The Mixtape** (Cunningham) - Free online
- **LOST** (Library of Statistical Techniques) - Code examples

### Software Documentation
- `did` package: [bcallaway11.github.io/did/](https://bcallaway11.github.io/did/)
- `fixest`: [lrberge.github.io/fixest/](https://lrberge.github.io/fixest/)

## Questions & Discussion {background-color="#2C5F8D"}

::: {.columns}
::: {.column width="50%"}
### Common Questions

- When should I use DiD vs RCT?
- How many pre-periods do I need?
- What if I only have 2 time periods?
- Can I use continuous treatments?
:::

::: {.column width="50%"}
### Get Started

- Install the `did` package
- Try examples with your data
- Check pre-trends first
- Compare TWFE vs robust methods
:::
:::

::: {.callout-tip}
**Remember:** The goal is credible causal inference. Use modern methods, be transparent about assumptions, and report robustness checks.
:::

# Thank You! {background-color="#2C5F8D"}

::: {.r-fit-text}
Questions?
:::