---
title: "data.table stuff"
author: "Tyson S. Barrett"
date: "7/12/2019"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

The `data.table` package has been a fun thing to learn about this week. Since getting into R several years ago, I had known `data.table` was a fast and efficient tool for larger data sets. I had tried using it but got confused early on and, unfortunately, switched to what I knew (just `data.frame`s). 

Then came along the `tidyverse`. Suddenly, R was straightforward; I was making less errors, producing meaningful insights, and more fully enjoying data analysis. It was a great way to become more productive and transparent in my code.

Recently, though, I've come to realize that I had a renewed interest in `data.table`. But it is difficult to break out of the habit of the same workflow I had been using for the past few years with `dplyr` and friends. But then I came across Elio's [post](https://eliocamp.github.io/codigo-r/en/2019/07/why-i-love-data-table/) about using pipes (`%>%`) with `data.table`. This was incredibly eye-opening to me! I had no idea that the tools that I liked the most would work so flawlessly with `data.table`.

Since then I've been trying to learn as much as I could about what makes `data.table` great. Turns out, there are a lot! But with it, comes some learning, particularly about the ins-and-outs of "modifying by reference".

To help both you and me remember some of the ways that `data.table` works, I put this small post together.

## What does it mean to modify by reference?

Well, it essentially means we are not creating a new object when we make certain types of changes to a data table.

To start, let's create a data table.

```{r}
library(magrittr)
library(data.table)
library(dplyr)

dt <- data.table(x = rnorm(1000),
                 y = rnorm(1000))
dt
```

This object is, to the computer:

```{r}
tracemem(dt)
```

If we make changes to `dt` in the way `data.table` likes to, then the object is not copied (or else the criptic `"<...>"` message would show up telling us we made a copy). 

```{r}
dt[, z := rnorm(1000)]
dt
```

This approach means that if we make a new pointer (in essence a new name that points to the same data), then changes to one makes changes to the other. This can be seen with us creating a new "object" called `dt2`. In reality, `dt2` just points to the same data that `dt` was pointing to. So, if we change `dt` by reference, it will change `dt2` as well. (Note, by using `:= NULL`, we are removing that variable.)

```{r}
dt2 <- dt
dt[, z := NULL]

dt2
```

This is different than what happens when we don't modify by reference. Notice that below, we create and modify the `dt3` object (originally it was pointing to the same data as `dt`). At the moment we make a change, a new data object is "created" (thus the tracemem output telling us we made a copy).

```{r}
dt3 <- dt
dt3$a <- rnorm(1000)
dt3
```

We can also see this by looking at our `dt` object and making sure there isn't a new variable called `a` in it.

```{r}
dt
```

So why would `data.table` want to perform this way? It is because it is far more efficient. We aren't making copies of things every time we make a change. This is hugely beneficial in large data sets, where RAM can get used up with all the copies. In smaller data, this is often not a big issue. 

Notably, the `dt3$a <- rnorm(1000)` code doesn't change the type of object it is---its still a data table. That means, you have the choice to use "by reference" or "by copy".

<div>
The take-home message is two-fold:

1. Modifying by reference is efficient.
2. `data.table` gives you the option to modify by reference or by making a copy.

</div>

For more information, run:

`vignette("datatable-reference-semantics")`

in R to see the great vignette explaining this further (and more clearly).

## So what happens when we pipe?

As I mentioned before, it was mind-blowing to see piping with `data.table`. Several have commented and said this probably slows it down a great deal. Again, Elio provided insight in this tweet:

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">A quick benchmark suggest that using pipes to chain data.table operations doesn&#39;t add any appreciable overhead. <a href="https://twitter.com/hashtag/rstats?src=hash&amp;ref_src=twsrc%5Etfw">#rstats</a><a href="https://t.co/zp0ac7tymn">https://t.co/zp0ac7tymn</a> <a href="https://t.co/pSkajjglEQ">pic.twitter.com/pSkajjglEQ</a></p>&mdash; Elio Campitelli (@d_olivaw) <a href="https://twitter.com/d_olivaw/status/1149029823567863808?ref_src=twsrc%5Etfw">July 10, 2019</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script> 

So, nope. Piping is not a slower version of using `data.table`. Rather, it is about using the fast features of `data.table` that will give you the benefit, not how you pipe/chain.

To see how this ends up working, consider:

```{r}
dt[, z := rnorm(1000)] %>% 
  .[, binary_z := case_when(z > 0 ~ 1, 
                            TRUE ~ 0)] %>% 
  .[, x_y := x + y]
```

Note that the `tracemem` message didn't show up. So what does that mean? It modified by reference, even with the pipe. Also, we used a `dplyr` function (`case_when()`) inside of a data table. Is that legal?! Turns out, yes. Yes, it is. As long as the `dplyr` function works on a vector, you can go ahead and use it in `data.table`.

## So what is the point?

The point, here, is that `data.table` is a powerful framework for working with data. It does some things differently, and these things are important. If you ignore "by reference" changes, you'll probably regret it. But in the right situations (probably a lot of them), it is amazing to use.

<iframe src="https://giphy.com/embed/sDjIG2QtbXKta" width="451" height="480" frameBorder="0" class="giphy-embed" allowFullScreen></iframe><p><a href="https://giphy.com/gifs/ron-swanson-parks-and-recreation-sDjIG2QtbXKta">via GIPHY</a></p>


{% if page.comments %}
<script>
    /**
     *  RECOMMENDED CONFIGURATION VARIABLES: EDIT AND UNCOMMENT THE SECTION BELOW TO INSERT DYNAMIC VALUES FROM YOUR PLATFORM OR CMS.
     *  LEARN WHY DEFINING THESE VARIABLES IS IMPORTANT: https://disqus.com/admin/universalcode/#configuration-variables
     */
    /*
    var disqus_config = function () {
        this.page.url = page.url;  // Replace PAGE_URL with your page's canonical URL variable
        this.page.identifier = page.identifer; // Replace PAGE_IDENTIFIER with your page's unique identifier variable
    };
    */
    (function() {  // DON'T EDIT BELOW THIS LINE
        var d = document, s = d.createElement('script');
        
        s.src = '//tysonstanley.disqus.com/embed.js';
        
        s.setAttribute('data-timestamp', +new Date());
        (d.head || d.body).appendChild(s);
    })();
</script>
<noscript>
Please enable JavaScript to view the
<a href="https://disqus.com/?ref_noscript" rel="nofollow">comments
powered by Disqus.</a>
</noscript>
{% endif %}








